---
title: "HR_Analytics_Data"
format: html
editor: visual
---

# HR Analytics Project – Synthetic HR Datasets

This project creates synthetic HR-related datasets to simulate common business questions in people analytics, such as pay equity, employee engagement, turnover prediction, recruiting source effectiveness, performance evaluation, and workplace safety.

The generated CSV files can be used for: - Exploratory HR data analysis\
- Predictive modeling (e.g., churn/attrition prediction)\
- Dashboard creation in R Shiny, Tableau, or Power BI\
- Portfolio projects to showcase HR analytics capabilities

# Recruitment data

(Hiring & Sales Quota)

-   Variables:

    -   attrition (Yes/No)

    -   performance_rating (1-5)

    -   sales_quota_pct (0.3 = 30% to 2.5 = 250% of target)

    -   recruiting_source

```{r}

set.seed(105)
n <- 500
attrition <- sample(c("Yes","No"), n, TRUE)
performance_rating <- sample(1:5, n, TRUE)
sales_quota_pct <- round(runif(n, min = 0.3, max = 2.5), 3)
recruiting_source <- sample(c("Applied Online","Campus","Referral ","Search Firm","N/A"), n, TRUE)

recruitment_data <- data.frame(attrition, performance_rating, sales_quota_pct, recruiting_source)
head(recruitment_data)
tail(recruitment_data)
write.csv(recruitment_data, "recruitment_data.csv", row.names = FALSE)

```

# Fair pay data

(Compensation & Job Level)

-   Variables:

    -   employee_id

    -   department

    -   salary

    -   new_hire (Yes/No)

    -    job_level (Hourly, Salaried, Manager)

```{r}
# ---- Fair Pay Data ----
set.seed(104)
n <- 2000
employee_id <- 1:n
department <- sample(c("HR","IT","Finance","Operations","Marketing","Sales"), n, TRUE)
salary <- round(rnorm(n, mean=70000, sd=15000), -2)
new_hire <- sample(c("Yes","No"), n, TRUE, prob = c(0.3,0.7))
job_level <- sample(c("Hourly","Salaried","Manager"), 
                    n, 
                    replace = TRUE)

fair_pay <- data.frame(employee_id, department, salary, new_hire, job_level)
head(fair_pay)
tail(fair_pay)

write.csv(fair_pay, "fair_pay_data.csv", row.names = FALSE)

```

# Performance review data

(Annual Ratings)

-   Variables:

    -   employee_id

    -   rating (1-5)

```{r}
set.seed(106)
n <- 2000
employee_id <- 1:n
rating <- sample(1:5, n, TRUE)

performance_data <- data.frame(employee_id, rating)
head(performance_data)
tail(performance_data)
write.csv(performance_data, "performance_data.csv", row.names = FALSE)

```

# accident data

```{r}
set.seed(103)
n <- 550
year <- sample(2018:2024, n, TRUE)
employee_id <- 1:n
accident_type <- sample(c("Mild","Moderate","Severe"), n, TRUE)

accident_data <- data.frame(year, employee_id, accident_type)
head(accident_data)
tail(accident_data)
write.csv(accident_data, "accident_data.csv", row.names = FALSE)

```

# survey data 2

(Yearly Engagement Tracking)

-   Variables:

    -   year

    -   employee_id

    -   engagement (1-5 scale)

```{r}
set.seed(102)
n <- 2000
year <- sample(2018:2024, n, replace = TRUE)
employee_id <- 1:n
engagement <- sample(1:5, n, replace = TRUE)

survey_data_2 <- data.frame(year, employee_id, engagement)
head(survey_data_2)
tail(survey_data_2)
write.csv(survey_data_2, "survey_data_2.csv", row.names = FALSE)

```

# survey data

(Employee Satisfaction & Benefits Usage)

-   Variables:

    -   employee_id

    -   department

    -   engagement (1-5 scale)

    -   salary (annual)

    -   vacation (days allocated)

    -   days_taken (days of vacation used)

```{r}
set.seed(101)
n <- 3000
employee_id <- 1:n
department <- sample(c("Engineering", "HR","Finance","IT","Operations","Marketing","Sales"), n, TRUE)
engagement <- sample(1:5, n, replace = TRUE)
salary <- round(rnorm(n, mean=70000, sd=15000), -2)
vacation <- sample(10:30, n, TRUE)
Vacation_days_taken <- sample(0:30, n, TRUE)

survey_data <- data.frame(employee_id, department, engagement, salary, vacation, Vacation_days_taken)
head(survey_data)
tail(survey_data)
write.csv(survey_data, "survey_data.csv", row.names = FALSE)

```

# HR data 2

(Overtime & Location)

-   Variables:

    -   year

    -   employee_id

    -   location

    -   overtime_hours

```{r}
set.seed(107)
n <- 4000
year <- sample(2018:2024, n, TRUE)
employee_id <- 1:n
location <- sample(c("NY","TX","CA","FL","Remote"), n, TRUE)
overtime_hours <- sample(0:50, n, TRUE)

hr_data_2 <- data.frame(year, employee_id, location, overtime_hours)
head(hr_data_2)
tail(hr_data_2)
write.csv(hr_data_2, "hr_data_2.csv", row.names = FALSE)
```

# HR data

(Workforce Demographics)

-   Variables:

    -   employee_id

    -   department

    -   job_level (Hourly, Salaried, Manager)

    -   gender

```{r}
set.seed(108)
n <- 2000
employee_id <- 1:n
department <- sample(c("Engineering","Finance","IT","Operations","Marketing","Sales"), n, TRUE)
job_level <- sample(c("Hourly","Salaried","Manager"), 
                    n, 
                    replace = TRUE)
gender <- sample(c("Male","Female","Other"), n, TRUE)

hr_data <- data.frame(employee_id, department, job_level, gender)
head(hr_data)
tail(hr_data)
write.csv(hr_data, "hr_data.csv", row.names = FALSE)
```
